Measuring causal influence with\\ back-to-back regression: the linear case

### 2019-05-23 paper/introduction.tex
refs

First, causal and non-causal factors may be numerous and collinear. In physics, for example, one may be challenged to identify whether fusion is caused by a change in temperature or in pressure, as these two factors can be difficult to manipulate independently. This issue becomes increasingly pronounced as the number of potential factor increases. In neuroscience, for example, it can be challenging to identify whether the frequency of a word presented on a subject's retina modulates brain activity as this factor covaries with numerous other factors such as word length (short words are more frequent than long words) word category (determinants are more frequent than adverbs), individual letters etc (e.g. words that contain "z" are rare) \citep{kutas2011thirty,pegado2014timing}. Generating a set of words that control for all of these factors simultaneously can be challenging; it is thus common to use forward modeling, i.e. to train a model to predict observations (e.g. brain activity) from a minimal combination of competing factors (e.g. word length, word frequency, etc, e.g. \citep{huth2016natural}), and investigate the resulting parameters of the best model.

@@

Both forward and backward modeling have competing benefits and drawbacks. Specifically, forward modeling disentangles the independent contribution of collinear factors, but does not combine multidimensional observations. By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \citep{weichwald2015causal, hebart_neuroimage_2015, king2018encoding}.

>>>>

First, causal and non-causal factors may be numerous and collinear. In physics, for example, one may be challenged to identify whether fusion is caused by a change in temperature or in pressure, as these two factors can be difficult to manipulate independently. This issue becomes increasingly pronounced as the number of potential factor increases. In neuroscience, for example, it can be challenging to identify whether the frequency of a word presented on a subject's retina modulates brain activity as this factor covaries with numerous other factors such as word length (short words are more frequent than long words) word category (determinants are more frequent than adverbs), individual letters etc (e.g. words that contain "z" are rare) \citep{kutas2011thirty,pegado2014timing}. Generating a set of words that control for all of these factors simultaneously can be challenging; it is thus common to use forward modeling, i.e. to train a model to predict observations (e.g. brain activity) from a minimal combination of competing factors (e.g. word length, word frequency, e.g. \citep{huth2016natural}), and investigate the resulting parameters of the best model.

@@

Both forward and backward modeling have competing benefits and drawbacks. Specifically, forward modeling disentangles the independent contribution of collinear factors, but does not combine multidimensional observations. By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \citep{weichwald2015causal, hebart2018deconstructing, king2018encoding}.

### 2019-05-23 paper/paper.tex
meg
refs

\title{Measuring causal influence with\\ back-to-back regression}

@@

\begin{abstract}
Identifying causes from observations is at the core of science. This endeavor
is particularly challenging when i) potential factors are difficult to
manipulate individually and ii) observations are complex and multi-dimensional.
To address this issue, we introduce ``Back-to-Back'' regression (B2B), a
method designed to measure, from a set of co-varying factors, the causal influences
that most plausibly account for multidimensional observations. After proving the consistency of B2B, we show that our method outperforms least-squares and related
regression and cross-decomposition  techniques (e.g. canonical
correlation analysis and partial least squares) on two tasks: causal
identification and out-of-distribution prediction. Finally, we apply B2B to
neuroimaging recordings of 102 subjects reading word sequences. The results
show that the early and late brain responses caused by low- and high-level
word features are more easily detected with B2B than with standard forward
regression whereas techniques.
% conclude  open
\end{abstract}

@@

    \node[] (labY) at (\posY, 3){Observations};
    \node[] (labX) at (\posX, 3){Factors};
    \node[] (labE) at (\posE, 3.3){Cause};
    \node[] (labE) at (\posE, 3){selection};
    \node[] (labN) at (\posN, 3){Noise};
    \node[] (labF) at (\posF, 3.3){Cause-effect};
    \node[] (labF) at (\posF, 3){mapping};

    \node[] (labY) at (\posY, 2)   {\color{gray} MEG recordings};
    \node[] (labX) at (\posX, 2)   {\color{gray} stimulus};
    \node[] (labX) at (\posX, 2.3) {\color{gray} features};
    \node[] (labE) at (\posE, 2.3) {\color{gray} causal};
    \node[] (labE) at (\posE, 2)   {\color{gray} features};
    \node[] (labN) at (\posN, 2)   {\color{gray} breathing\ldots};
    \node[] (labF) at (\posN, 2.3) {\color{gray} bloodflow,};
    \node[] (labF) at (\posF, 2)   {\color{gray} MEG helmet};

>>>>

\title{Measuring causal influence with\\ back-to-back regression: the linear case}

@@

\begin{abstract}
Identifying causes from observations is at the core of science. This endeavor
is particularly challenging when i) potential factors are difficult to
manipulate individually and ii) observations are complex and multi-dimensional.
To address this issue, we introduce ``Back-to-Back'' regression (B2B), a
method designed to measure, from a set of co-varying factors, the causal
influences that most plausibly account for multidimensional observations. After
proving the consistency of B2B, we show that our method outperforms
least-squares and related regression and cross-decomposition techniques (e.g.
canonical correlation analysis and partial least squares) on two tasks: causal
identification and out-of-distribution prediction. Finally, we apply B2B to
neuroimaging recordings of 102 subjects reading word sequences. The results
show that the early and late brain responses caused by low- and high-level
word features respectively are more easily detected with B2B than with standard forward regression.
% conclude  open
\end{abstract}

@@

    \node[] (labY) at (\posY, 2){Observations};
    \node[] (labX) at (\posX, 2){Factors};
    \node[] (labE) at (\posE, 2){Cause};
    \node[] (labE) at (\posE, 1.7){selection};
    \node[] (labN) at (\posN, 2){Noise};
    \node[] (labF) at (\posF, 2){Cause-effect};
    \node[] (labF) at (\posF, 1.7){mapping};

    % \node[] (labY) at (\posY, 2)   {\color{gray} MEG recordings};
    % \node[] (labX) at (\posX, 2)   {\color{gray} stimulus};
    % \node[] (labX) at (\posX, 2.3) {\color{gray} features};
    % \node[] (labE) at (\posE, 2.3) {\color{gray} causal};
    % \node[] (labE) at (\posE, 2)   {\color{gray} features};
    % \node[] (labN) at (\posN, 2)   {\color{gray} breathing\ldots};
    % \node[] (labF) at (\posN, 2.3) {\color{gray} bloodflow,};
    % \node[] (labF) at (\posF, 2)   {\color{gray} MEG helmet};
